---
title: 'Model-agnostic methods: Partial Dependency Plot (PDP)'
author: "Natalia Martínez, Elena Orón, Aitana Sebastià"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

```{r}
# setwd("C:/Users/natal/Downloads/EDM/PL Tema 5/Practica 5/")
setwd("C:/Users/34686/3CD - 23_24/EDM/XAI3")
```

# Libraries

```{r}

# library(randomForest)

# Data preparation
library(dplyr)

# Data partition
library(caret)  

# PDP
library(pdp)
library(ggplot2)

```

# 1. Bike-rental database

## Data

```{r}
day <- read.csv("day.csv")
hour <- read.csv("hour.csv")
```

## Data preprocessing

```{r}
## We prepare the new variables

# We just include 3 out of the 4 features to undo the correlation
day$season = factor(day$season)
season_one_hot = model.matrix(~ 0 + season, data = day)

# One-hot encoding
misty = ifelse(day$weathersit==2,1,0)
rain = ifelse(day$weathersit %in% c(3,4),1,0)

# Unnormalize: temp -> (t-t_min)/(t_max-t_min), t_min=-8, t_max=+39 (only in hourly scale) 
temp = (39+8)*day$temp -8
  
# Unnormalize: hum -> The values are divided to 100 (max) 
hum = day$hum*100
  
# Unnormalize: windspeed -> The values are divided to 67 (max)
windspeed = day$windspeed*67

# length(day$dteday) --> 731 days
days_since_2011 =  seq(0,365*2)
dim(day)
```

```{r}
day_new = data.frame(day$cnt, day$workingday, day$holiday, season_one_hot[,-1], misty, rain, temp, hum, windspeed, days_since_2011)
names(day_new)[1:6] = c("cnt", "workingday", "holiday", "spring", "summer", "fall")
day_new
```

The partial dependence plot shows the marginal effect of a feature on the predicted outcome of a previously fit model.

Apply PDP to the regression example of predicting bike rentals. Fit a random forest approximation for the prediction of bike rentals (cnt). Use the partial dependence plot to visualize the relationships the model learned. Use the slides shown in class as model.

Generate a 2D Partial Dependency Plot with humidity and temperature to predict the number of bikes rented depending on those parameters.

BE CAREFUL: due to the size, extract a set of random samples from the BBDD before generating the data for the Partial Dependency Plot.Show the density distribution of both input features with the 2D plot as shown in the class slides.\
TIP: Use geom_tile() to generate the 2D plot. Set width and height to avoid holes.

## Random Forest

```{r}
set.seed(123)

# Using randomForest library
# rf <- randomForest(formula = cnt ~ ., data = day_new)

# Train and test split
index <- createDataPartition(day_new$cnt, p = .8, list = FALSE)
day_train <- day_new[ index, ]
day_test <- day_new[-index, ]

# Model building
rf_day <- train(cnt ~ . , data = day_train, method = "rf")

```

## One dimensional Partial Dependence Plot

General function to plot the required variables.

```{r}

# Define the function to generate the PDP plot
generate_pdp_plot <- function(model, pred_var, var_name, train_data, y_name="Predicted bike rentals", grid_resolution = 20) {
  # Generate the PDP data for the specified variable
  pdp_data <- partial(model, pred.var = pred_var, train = train_data, grid.resolution = grid_resolution)
  
  # Convert the PDP data to a data frame
  pdp_df <- as.data.frame(pdp_data)
  
  # Generate the PDP plot
  pdp_plot <- ggplot(pdp_df) +
    geom_line(aes_string(x = pred_var, y = "yhat"), color = "black") +  # Line for the PDP
    geom_rug(data = train_data, aes_string(x = pred_var), sides = "b") +
    theme_minimal() +
    labs(title = paste("Partial dependence plot for", var_name),
         x = pred_var, y = y_name)
  
  # Return the PDP plot
  return(pdp_plot)
}
```

Plot the PDP for numerical variables:

```{r}
pdp_plot_hum <- generate_pdp_plot(rf_day, "hum", "Humidity" ,day_new)
print(pdp_plot_hum)

pdp_plot_temp <- generate_pdp_plot(rf_day, "temp", "Temperature" ,day_new)
print(pdp_plot_temp)

pdp_plot_windspeed <- generate_pdp_plot(rf_day, "windspeed", "Wind speed",day_new)
print(pdp_plot_windspeed)

pdp_plot_days_since_2011 <- generate_pdp_plot(rf_day, "days_since_2011", "Days since 2011",day_new)
print(pdp_plot_days_since_2011)


```

## Bidimensional Partial Dependence Plot

```{r}
# Generate PDP data for humidity and temperature
pdp_data <- partial(rf_day, pred.var = c("hum", "temp"), train = day_new, grid.resolution = 20)

# Convert the pdp_data to a data frame
pdp_df <- as.data.frame(pdp_data)

# Normalize yhat values between 0 and 1
#pdp_df$yhat <- (pdp_df$yhat - min(pdp_df$yhat)) / (max(pdp_df$yhat) - min(pdp_df$yhat))


# Generate the 2D plot using geom_tile()
pdp_2d_plot <- ggplot(pdp_df, aes(x = hum, y = temp, fill = yhat)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "2D Partial Dependency Plot",
       x = "Humidity", y = "Temperature", fill = "Predicted Bike Rentals")

# Display the plot
ggplot() +
  geom_tile(data = pdp_df, aes(x = hum, y = temp, fill = yhat), width = 10, height = 5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  geom_rug(data = day_new, aes(x = hum), sides = "b") +
  geom_rug(data = day_new, aes(y = temp), sides = "l") + 
  theme_minimal()

```

# 2. House price database

Apply the previous concepts to predict the price of a house from the database kc_house_data.csv. In this case, use again a random forest approximation for the prediction based on the features bedrooms, bathrooms, sqft_living, sqft_lot, floors and yr_built. Use the partial dependence plot to visualize the relationships the model learned.

BE CAREFUL: due to the size, extract a set of random samples from the BBDD before generating the data for the Partial Dependency Plot.

## Data

```{r}
house = read.csv("kc_house_data.csv")
```

## Data preparation

```{r}
house_new = select(house,price, bedrooms, bathrooms, sqft_living, sqft_lot, floors, yr_built)

# We sample the data to obtain the 25% of it
house_sample = sample_frac(house_new, 0.25)
```

## Random forest

```{r}
set.seed(123)
# Train and test split
index <- createDataPartition(house_sample$price, p = .8, list = FALSE)
house_train <- house_sample[ index, ]
house_test <- house_sample[-index, ]

# Model building
rf_house <- train(price ~ . , data = house_train, method = "rf")
pred_rf_house <- predict(rf_house, house_test)

```

## One dimensional Partial Dependence Plot



```{r}
pdp_plot_bedrooms <- generate_pdp_plot(rf_house, "bedrooms", "Bedrooms" ,house_sample, "Predicted Price")
print(pdp_plot_bedrooms)

pdp_plot_bathrooms <- generate_pdp_plot(rf_house, "bathrooms", "Bathrooms" ,house_sample, "Predicted Price")
print(pdp_plot_bathrooms)

pdp_plot_sqft_living <- generate_pdp_plot(rf_house, "sqft_living", "SQFT living",house_sample, "Predicted Price")
print(pdp_plot_sqft_living)

pdp_plot_floors <- generate_pdp_plot(rf_house, "floors", "Floors", house_sample, "Predicted Price")
print(pdp_plot_floors)
```


```{r}
summary(house_sample)
```


## Bidimensional Partial Dependence Plot

```{r}

```
